Exploring mathematical objects from custom-tailored mathematical universes

### 2020-06-21 paper-filmat.tex
Some additions (in response to review)

\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
multiverse view in set theory~\cite{hamkins:multiverse}. In fact, the
topos-theoretic multiverse can be regarded as an extension of the set-theoretic
multiverse: While Hamkins proposes to embrace all models of set theory (not
necessarily all of them not equally), we propose to embrace all toposes (again
not necessarily all of them not equally). As every model~$M$ of set theory
gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
topos-theoretic one.

@@

\begin{ex}\textbf{``Heyting arithmetic is categorical.''} In addition to the standard
model~$\NN$, the standard topos contains uncountably many nonstandard models of
Peano arithmetic (at least if we assume a classical meta level). By a theorem
of Benno van den Berg and Jaap van Oosten, the situation
is quite different in the \effective topos~\cite{berg-oosten:arithmetic}:
\begin{enumerate}
\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, has
exactly one model up to isomorphism, namely~$\NN$.
\item In fact, even the finitely axiomatizable subsystem of Heyting arithmetic
where the induction scheme is restricted to~$\Sigma_1$-formulas has exactly one
model up to isomorphism, again~$\NN$. As a consequence, Heyting arithmetic is
finitely axiomatizable.
\item Peano arithmetic is ``quasi-inconsistent'' in that it does not have any
models, for any model of Peano arithmetic would also be a model of Heyting
arithmetic, but the only model of Heyting arithmetic is~$\NN$ and~$\NN$ does
not validate the theorem ``any Turing machine terminates or does not
terminate'' of Peano arithmetic.
\end{enumerate}
As a consequence, Gödel's completeness theorem fails in the \effective topos:
In the \effective topos, Peano arithmetic is consistent (because it is
equiconsistent to Heyting arithmetic, which has a model) but does not have a
model.

@@

A small variant of this lens has been used to give a new proof of
\emph{Grothendieck's generic freeness lemma}, a fundamental theorem in
algebraic geometry. The new proof uses the lens to reduce to the case
of fields, where the claim is trivial~\cite[Section~11.5]{blechschmidt:phd}.
\end{rem}

Within~$\Zar(\RR)$, we can construct the set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ validates the following laws:
\begin{enumerate}
\item Law of cancellation: $\forall x \? \RR^\sim\_ \forall y \? \RR^\sim\_ \bigl((\forall
\varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
\Rightarrow x = y\bigr)$
\item Axiom of micro-affinity: $\forall f \? (\RR^\sim)^\Delta\_ \exists! a \? \RR^\sim\_
\forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
\end{enumerate}
The unique number~$a$ in the axiom of micro-affinity deserves to be
called~``$f'(0)$''; this is how we synthetically define the derivative in
synthetic differential geometry.

@@

This property is where the language of toposes derives its usefulness for
mathematical practice from: In some cases, the easiest way to prove a result
(about objects of the standard topos) is
\begin{enumerate}
\item to observe that the claim is
equivalent to the Kripke--Joyal translation of a different (typically more
perspicuous) claim about objects of some problem-specific relevant topos and
then
\item to verify this different claim, reasoning internally to the topos.
\end{enumerate}

@@

% FUTURE: Cite blechschmidt:generic-freeness?

>>>>

\begin{rem}The picture of a topos-theoretic multiverse is related to Hamkin's
multiverse view in set theory~\cite{hamkins:multiverse}. In fact, the
topos-theoretic multiverse can be regarded as an extension of the set-theoretic
multiverse: While Hamkins proposes to embrace all models of set theory (not
necessarily all of them equally -- we might prefer some models over others), we propose to embrace all toposes (again
not necessarily all of them equally). As every model~$M$ of set theory
gives rise to a topos~$\Set_M$, the set-theoretic multiverse is contained in the
topos-theoretic one.

@@

\begin{ex}\textbf{``Heyting arithmetic is categorical.''} In addition to the standard
model~$\NN$, the standard topos contains uncountably many nonstandard models of
Peano arithmetic (at least if we assume a classical meta level). By a theorem
of Benno van den Berg and Jaap van Oosten, the situation
is quite different in the \effective topos~\cite{berg-oosten:arithmetic}:
\begin{enumerate}
\item Heyting arithmetic, the intuitionistic cousin of Peano arithmetic, is
categorical in the sense that it has
exactly one model up to isomorphism, namely~$\NN$.
\item In fact, even the finitely axiomatizable subsystem of Heyting arithmetic
where the induction scheme is restricted to~$\Sigma_1$-formulas has exactly one
model up to isomorphism, again~$\NN$. As a consequence, Heyting arithmetic is
finitely axiomatizable.
\item Peano arithmetic is ``quasi-inconsistent'' in that it does not have any
models, for any model of Peano arithmetic would also be a model of Heyting
arithmetic, but the only model of Heyting arithmetic is~$\NN$ and~$\NN$ does
not validate the theorem ``any Turing machine terminates or does not
terminate'' of Peano arithmetic.
\end{enumerate}
As a consequence, Gödel's completeness theorem fails in the \effective topos:
In the \effective topos, Peano arithmetic is consistent (because it is
equiconsistent to Heyting arithmetic, which has a model) but does not have a
model.

@@

A small variant of this lens has been used to give a new proof of
\emph{Grothendieck's generic freeness lemma}, a fundamental theorem in
algebraic geometry about the free locus of certain sheaves. The new proof uses the lens to reduce to the case
of fields, where the claim is trivial~\cite[Section~11.5]{blechschmidt:phd},
and improves in length on all previously known proofs, even if the topos
machinery is eliminated by unrolling the appropriate
definitions~\cite{blechschmidt:generic-freeness}.\footnote{This contribution is
not the proper place for an exposition of Grothendieck's generic freeness
lemma, but some aspects can already be appreciated on a syntactical level.
Grothendieck's generic freeness lemma states that any finitely generated sheaf
of modules on a reduced scheme is finite locally free \emph{on a dense open}.
By employing the internal language, this statement is reduced to the following
fact of intuitionistic linear algebra: Any finitely generated module over a
field is \notnot finite free.}
\end{rem}

Within~$\Zar(\RR)$, we can construct the set~$\Delta \defeq \{ \varepsilon \? \RR^\sim \,|\,
\varepsilon^2 = 0 \}$ of nilsquare numbers. Then~$\RR^\sim$ validates the following laws:
\begin{enumerate}
\item Law of cancellation: $\forall x \? \RR^\sim\_ \forall y \? \RR^\sim\_ \bigl((\forall
\varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
\Rightarrow x = y\bigr)$
\item Axiom of micro-affinity: $\forall f \? (\RR^\sim)^\Delta\_ \exists! a \? \RR^\sim\_
\forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
\end{enumerate}
The unique number~$a$ in the axiom of micro-affinity deserves to be
called~``$f'(0)$''; this is how we synthetically define the derivative in
synthetic differential geometry. (However, despite these properties the Zariski
topos is not yet \emph{well-adapted} to synthetic differential geometry in the
sense of Definition~\ref{defn:well-adapted} below.)

@@

This property is where the language of toposes derives its usefulness for
mathematical practice from: In some cases, the easiest way to prove a result
(about objects of the standard topos) is
\begin{enumerate}
\item to observe that the claim is
equivalent to the Kripke--Joyal translation of a different (typically more
transparent) claim about objects of some problem-specific relevant topos and
then
\item to verify this different claim, reasoning internally to the topos.
\end{enumerate}

@@
